---
title: "Making Maps in R"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

| *AnEx Learning Summit*
| *Thursday October 21, 2021*
| 
| *Sarah Rankin*
| *Quantitative Data Analyst*
| *Stategy and Public Impact*
| *New York Public Library*
| *sarahrankin\@nypl.org*
| 

## Topics

-   Static maps with sf and ggplot
-   Dynamic maps with leaflet

## Setup

This is an Rmarkdown notebook. If you're viewing it in a web browser, you can click on Code-\>Download Rmd in the top right corner, then open it in RStudio and to run/edit the code. If you don't use RStudio you can also just copy the code into your editor of choice.

We'll be using a variety of `tidyverse` packages as well as `sf` and `leaflet`. If you don't have these packages installed, uncomment the following code and install them now. (It'll take a little while.)

```{r}
# install.packages("tidyverse")
# install.packages("sf")
# install.packages("leaflet")

```

```{r message=FALSE, warning=FALSE}
#load just the tidyverse for now
library(tidyverse)

```

## Quick ggplot refresher

Let's create some very simple point data.

```{r}
five_points <- data.frame(var1 = 1:5, var2 = as.integer(c(3,5,4,4,1)))
five_points 
```

-   Call `ggplot()` to make a canvas

-   Define aesthetics with `aes()` - these map your data to spatial properties like the x and y dimensions, or visual aesthetics like color, fill, alpha etc

-   Add a geom - here, `geom_point()`

```{r}
five_points %>% 
  ggplot(aes(x=var1, y=var2)) +
  geom_point()
  
```

-   Add flags indicating whether each variable is even or odd (`%%` is the modulo function)

```{r}
five_points <- five_points %>%  
  mutate(var1_odd_or_even = ifelse(var1%%2==0,"even","odd"),
         var2_odd_or_even = ifelse(var2%%2==0,"even","odd"))

five_points


```

-   Map "var_1\_is even" to the color aesthetic

```{r}
five_points %>% 
  ggplot(aes(x=var1, y=var2,color = var1_odd_or_even)) +
  geom_point(size = 5) 


```

#### Aesthetics in ggplot

-   Color is for lines (including outlines of shapes)

-   Fill is for areas (filling in shapes)

    -   but our points are filled in based on the color aesthetic?!

    -   the default point shape is shape 19 ("circle") - just a solid circle with the color taken from the color aesthetic; shape 21 ("filled circle") is a circle with a fill and an outline, which can be specified separately

-   Generally, you can either map the aesthetic to a data point or specify its value outside of the `aes()` function. If you do neither, a default will be used.

-   Aesthetic mapping/specification can be done in the original ggplot() call, or within individual geoms

Good general reference on specifying aesthetics:

<https://ggplot2.tidyverse.org/articles/ggplot2-specs.html#point-1>

```{r}
five_points %>% 
  ggplot(aes(x=var1, 
             y=var2,
             color = var1_odd_or_even,
             fill = var2_odd_or_even)) +
  geom_point(size = 5, 
             shape = 21,
             stroke = 2) 

```

### Points as Maps

-   Longitude and Latitude are (on a 2d map) x and y coordinates

-   Let's get some coordinate data from the NYC facilities database

<https://data.cityofnewyork.us/City-Government/Facilities-Database/ji82-xba5>

```{r}
public_libraries <- jsonlite::fromJSON("https://data.cityofnewyork.us/resource/ji82-xba5.json?factype=PUBLIC%20LIBRARY") 

public_libraries %>% str()
```

We need latitude and longitude to be numeric, not strings

```{r}
public_libraries <- public_libraries %>%  
  mutate(latitude=as.numeric(latitude),longitude=as.numeric(longitude))

public_libraries %>% 
  select(facname,latitude,longitude) %>% 
  str()


```

-   Map the longitude and latitude to the x and y aesthetics

    -   Everyone has a strategy for remembering which is which - I like to think of soup dumplings - "**X**iao **long** bao" - credit to [\@seankross](https://twitter.com/seankross/status/1134326728476712960)

```{r}

public_libraries %>% 
  ggplot(aes(x=longitude,
             y=latitude)) + 
  geom_point() 



```

-   That does not look like a map. Let's look at our data.

```{r}
public_libraries %>% 
  select(latitude,longitude) %>% 
  summary()


```

-   At least some of our lat/long values are way out of expected range

```{r}
public_libraries %>% 
  filter(latitude<39|longitude>74) %>% 
  select(latitude,longitude,facname,addressnum,streetname,address,city,boro)


```

-   Looks like we can clean this up by just filtering out these two cases

```{r}

 public_libraries <- public_libraries %>%   
  filter(!latitude==0,
         !longitude==0)
 
 public_libraries %>% nrow()

```

-   Try again

```{r}
public_libraries %>% 
  ggplot(aes(x=longitude,
             y=latitude)) + 
  geom_point() 

```

-   Make it a bit prettier

    -   Color the points and rename the legend with `scale_color_manual()`

    -   Get rid of axes and background with `theme_void()`

```{r}
#define some system colors - a named vector of colors will map the colors to the values of the color aesthetic (backticks let you use non-syntactic R names)
#color values can be names from R's built-in colors or hex codes

library_system_colors <- c(`New York Public Library`="red3",
                   `Brooklyn Public Library` = "orange2",
                   `Queens Public Library` = "purple3")

public_libraries %>% 
  ggplot(aes(x=longitude,
             y=latitude,
             color = overagency)) + 
  geom_point() +
  scale_color_manual(values = library_system_colors,name = "System") +
  theme_void()



```

## The `sf` (simple features) package

<https://r-spatial.github.io/sf/>

```{r }
library(sf)
```

Simple features is a set of standards for defining two-dimensional geometries, used by various GIS systems, building up from x-y coordinates within a coordinate reference system (crs).

-   POINT

-   LINESTRING

-   POLYGON

-   MULTIPOINT

-   MULTIPOLYGON

`sf` package in R lets you:

-   store geometry data as a list-column in a data frame or tibble, alongside other data columns

-   import data into the R sf format from various GIS formats

-   manipulate sf geometries (concatenate, subtract, find overlaps, etc)

You can make an `sf` object directly from a data frame:

```{r}

five_points_sf <- five_points %>% st_as_sf(coords = c("var1","var2")) 

five_points_sf %>% print()
```

-   There is a special geom for sf objects: `geom_sf`

-   When you add geom_sf to a plot (and the `sf` package is loaded), ggplot will plot the geometry column - plotting method depends on the sf geometry type (point, polygon, etc)

```{r}

five_points_sf %>% 
  ggplot() + 
  geom_sf(
    aes(color = var1_odd_or_even,
              fill = var2_odd_or_even),
          size = 5,
          shape = 21,
          stroke = 2
    )


```

### Map Polygons

But for mapping we usually import geometry data from an external dataset.

Formats:

-   GEOJSON

-   ESRI shapefile

-   Many others (`st_drivers(what = "vector")`)

NYC geography resources:

<https://www1.nyc.gov/site/planning/data-maps/open-data/census-download-metadata.page>

-   Get the NYC PUMA (Public Use Microdata Area) geographies in GEOJSON format (PUMAs in NYC correspond - mostly - to Community Districts)

```{r cache=TRUE}

#read geojson format directly from download link
pumas_2010_geojson <- st_read("https://services5.arcgis.com/GfwWNkhOj9bNBqoJ/arcgis/rest/services/NYC_Public_Use_Microdata_Areas_PUMAs_2010/FeatureServer/0/query?where=1=1&outFields=*&outSR=4326&f=pgeojson") %>% 
  st_as_sf()

pumas_2010_geojson %>% print()

 


```

-   Shapefile format is also very common

```{r cache=TRUE}
#for shapefiles, download the zipped shapefile directory into your current working directory, then unzip (can do this outside R if you prefer!)
#download
download.file(url = "https://www1.nyc.gov/assets/planning/download/zip/data-maps/open-data/nypuma2010_21c.zip",destfile = "nypuma2010_21c.zip")
#unzip
unzip("nypuma2010_21c.zip")

#read in the shapefile
pumas_2010_shp <- st_read("nypuma2010_21c/nypuma2010.shp") %>% st_as_sf()


pumas_2010_shp %>% print()

```

-   Again, adding `geom_sf()` to a ggplot renders the data in the geometry column - in this case it draws polygons because the geometry type is multipolygon

```{r }
pumas_2010_geojson %>% 
  ggplot() + 
  geom_sf()

```

-   Add our points to this map by specifying different data sources within the `geom_sf` and `geom_point` calls

-   `sf` objects can have a CRS (coordinate reference system) defined

```{r }

  ggplot() + 
  geom_sf(data = pumas_2010_geojson,
    fill = "grey",
    color = "grey"
    ) +
  geom_point(data = public_libraries, 
             aes(x=longitude,
                 y=latitude,
                 color = overagency)) +
  scale_color_manual(name = "System",values = library_system_colors) +
  #adding coord sf ensures crs matches
  coord_sf() +
  theme_void()


```

### Chloropleths

-   Map areas colored by data value

#### Broadband adoption in New York City

-   NYC Open Data: [NYC's Internet Master Plan: Home Broadband and Mobile Broadband Adoption by PUMA](https://data.cityofnewyork.us/City-Government/Internet-Master-Plan-Home-Broadband-and-Mobile-Bro/g5ah-i2sh)

-   Based on American Community Survey Data; shows share of households that have both mobile and home broadband

```{r paged.print=FALSE, cache=TRUE}

broadband_use <- read_csv("https://data.cityofnewyork.us/api/views/g5ah-i2sh/rows.csv?accessType=DOWNLOAD")


#clean up the names a bit
broadband_use <- broadband_use %>% 
  rename(PUMA = `PUMA (Public Use Microdata Sample Areas)`,
         broadband_adoption = `Home Broadband and Mobile Broadband Adoption (Percentage of  Households)`,
         broadband_adoption_quartile = `Home Broadband and Mobile Broadband Adoption by Quartiles (High, Medium-High, Medium-Low, Low)`)

broadband_use %>% head()


```

-   Add the broadband data to our PUMA geom dataset

```{r cache=TRUE}

#convert PUMA to string; make adoption quartile into factor so it'll sort correctly  
broadband_use <- broadband_use %>% 
  mutate(PUMA = as.character(PUMA),
         broadband_adoption_quartile = factor(broadband_adoption_quartile,
                                              levels = c("High", "Medium High", "Medium Low", "Low")))

pumas_2010_geojson <- pumas_2010_geojson %>% 
  left_join(broadband_use, by = "PUMA")

pumas_2010_geojson %>% print()


```

-   Map the `geom_sf` fill aesthetic to broadband_adoption

-   Use scale_fill_gradient to specify fill colors

```{r}

  ggplot() + 
  geom_sf(data = pumas_2010_geojson,
    aes(fill = broadband_adoption),
    color = "grey"
    ) +
  geom_point(data = public_libraries,
             aes(x=longitude,
                 y=latitude,
                 color = overagency),
             size = 2) +
  scale_color_manual(name = "System",values = library_system_colors) +
  scale_fill_gradient(low = "grey70",high = "grey10", labels = scales::percent_format(accuracy = 1), name = "Broadband adoption") +
  coord_sf() +
  theme_void()

```

-   Or define discrete colors and use `scale_fill_manual`

```{r}

broadband_cols <- c("grey10","grey30","grey50","grey70") %>% set_names(levels(pumas_2010_geojson$broadband_adoption_quartile))


ggplot() + 
  geom_sf(data = pumas_2010_geojson,
    aes(fill = broadband_adoption_quartile),
    color = "grey"
    ) +
  geom_point(data = public_libraries,
             aes(x=longitude,
                 y=latitude,
                 color = overagency),
             size = 2) +
  scale_color_manual(name = "System",values = library_system_colors) +
  scale_fill_manual(values = broadband_cols, name = "Broadband adoption") +
  coord_sf() +
  theme_void()


```

#### Median income

<https://www1.nyc.gov/site/planning/planning-level/nyc-population/american-community-survey.page>

-   Download data and add it to our pumas_2010_geojson data frame

```{r cache=TRUE}


download.file(url = "https://www1.nyc.gov/assets/planning/download/office/planning-level/nyc-population/acs/econ_2018_acs5yr_puma.xlsx",
              destfile = "econ_2018_acs5yr_puma.xlsx")

econdata_puma <- readxl::read_xlsx("econ_2018_acs5yr_puma.xlsx", sheet = "EconData")

pumas_2010_geojson <- pumas_2010_geojson %>% 
  left_join(econdata_puma %>% 
              select(PUMA = GeoID,GeogName,median_household_income = MdHHIncE) %>% 
              mutate(median_income_category = 
                       cut(median_household_income,
                           breaks = c(0,50000,75000,100000,150000),
                           labels = c("$0-50K","$50-75K","$75-100K","$100K+"))))


```

-   Create a new color scale with some greens

```{r}
median_income_category_cols <- c("#a9ccbc","#7fb29b","#4c7f68","#335545") %>% set_names(levels(pumas_2010_geojson$median_income_category))

ggplot() + 
  geom_sf(data = pumas_2010_geojson,
    aes(fill = median_income_category),
    color = "grey"
    ) +
  geom_point(data = public_libraries,
             aes(x=longitude,
                 y=latitude,
                 color = overagency),
             size = 2) +
  scale_color_manual(name = "System",values = library_system_colors) +
  scale_fill_manual(values = median_income_category_cols, name = "Median income") +
  coord_sf() +
  theme_void() 

```

# Leaflet: interactive maps

```{r}
library(leaflet)

```

[Leaflet](https://leafletjs.com): open source javascript library for making interactive maps on various platforms

R has a `leaflet` package that lets you create leaflet "widgets" within R

Create a map, add tiles, set the view

-   Similar to ggplot in that we keep adding layers/elements, except we continue using the magrittr pipe `%>%` to add elements to a leaflet map (instead of the ggplot `+`)

-   for now we're creating a map without any data

```{r}
leaflet() %>% 
  #addTiles() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lat = 40.7,lng = -74, zoom = 10)

```

Add our public library branches

-   `addCircles()` adds points to the map

-   can specify the data source within the `addCircles()` call or in the original `leaflet()` call

-   to refer to fields in our data frame, use the tilde prefix \~

```{r}

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lat = 40.7,lng = -74, zoom = 10) %>%
  addCircles(data = public_libraries,
             lng = ~longitude,
             lat = ~latitude
             )


```

-   Use `colorFactor()` to create a palette to color the circles by system

-   Use the `label` and `popup` arguments to add hover/click info

```{r}
pal_system <- colorFactor(palette = library_system_colors ,
                          domain = unique(public_libraries$overagency),
                          ordered = T)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lat = 40.7,lng = -74, zoom = 10) %>%
  addCircles(data = public_libraries,
             lng = ~longitude,
             lat = ~latitude,
             color = ~pal_system(overagency),
             fill = ~pal_system(overagency),
             radius = 100,
             label = ~facname,
             popup = ~address
             ) %>% 
  addLegend(position = "topleft",
            pal = pal_system, 
            values = unique(public_libraries$overagency))




```

Because we have a basemap, we don't need the PUMA polygons to create a basic map. But we can add them if we want our chloropleths on the interactive map.

-   Fill color argument in leaflet is `fillColor;` specify opacity with `fillOpacity`

-   Use `stroke` to specify whether boundary lines are drawn; `color` , `weight`, and `opacity` to control their appearance

```{r}

pal_broadband <- colorFactor(palette = broadband_cols, 
                             domain = levels(pumas_2010_geojson$broadband_adoption_quartile),
                             ordered = T)


leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lat = 40.7,lng = -74, zoom = 10) %>%
  addPolygons(data = pumas_2010_geojson,
              fillColor = ~pal_broadband(broadband_adoption_quartile),
              fillOpacity = .7,
              color = "grey",
              stroke = T,
              weight = 1,
              label = ~PUMA,
              popup = ~paste0("<b>PUMA ",PUMA,"</b>",
                             "<br>", GeogName,
                             "<br>Broadband adoption: ",round(broadband_adoption*100,0),"%",
                             "<br>Broadband adoption category: ",broadband_adoption_quartile)) 


```

Combine it with the libraries and add a legend

-   Order matters - each successive layer is plotted on top of the previous ones

```{r}

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lat = 40.7,lng = -74, zoom = 10) %>%
  addPolygons(data = pumas_2010_geojson,
              fillColor = ~pal_broadband(broadband_adoption_quartile),
              fillOpacity = .7,
              color = "grey",
              stroke = T,
              weight = 1,
              label = ~PUMA,
              popup = ~paste0("<b>PUMA ",PUMA,"</b>",
                              "<br>", GeogName,
                             "<br>Broadband adoption: ",round(broadband_adoption*100,0),"%",
                             "<br>Broadband adoption category: ",broadband_adoption_quartile)) %>% 
    addCircles(data = public_libraries,
             lng = ~longitude,
             lat = ~latitude,
             color = ~pal_system(overagency),
             fill = ~pal_system(overagency),
             radius = 100,
             label = ~facname,
             popup = ~address
             ) %>% 
  addLegend(position = "topleft",
            pal = pal_system, 
            values = public_libraries$overagency,
            title = "Library System") %>% 
  addLegend(position = "topleft",
            pal = pal_broadband, 
            values =  pumas_2010_geojson$broadband_adoption_quartile,
            opacity = .7,
            title = "Broadband Access")

```

Visualize more than one layer

-   Use `group` argument within each layer to specify which group it belongs to

-   `addLayersControl` adds a control to toggle between layers or turn layers on and off

-   assigning the map to an object creates a leaflet object with class `htmlwidget`

```{r}
pal_income <- colorFactor(palette = median_income_category_cols, 
                             domain = levels(pumas_2010_geojson$median_income_category),
                             ordered = T)

libraries_map <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lat = 40.7,lng = -74, zoom = 10) %>%
  addPolygons(data = pumas_2010_geojson,
              fillColor = ~pal_broadband(broadband_adoption_quartile),
              fillOpacity = .7,
              color = "grey",
              stroke = T,
              weight = 1,
              label = ~PUMA,
              popup = ~paste0("<b>PUMA ",PUMA,"</b>",
                              "<br>", GeogName,
                             "<br>Broadband adoption: ",round(broadband_adoption*100,0),"%",
                             "<br>Broadband adoption category: ",broadband_adoption_quartile),
              group = "Broadband") %>% 
  addPolygons(data = pumas_2010_geojson,
              fillColor = ~pal_income(median_income_category),
              fillOpacity = .7,
              color = "grey",
              stroke = T,
              weight = 1,
              label = ~PUMA,
              popup = ~paste0("<b>PUMA ",PUMA,"</b>",
                              "<br>", GeogName,
                             "<br>Median income: $",format(median_household_income,big.mark = ","),
                             "<br>Median income category: ",median_income_category),
              group = "income") %>% 
    addCircles(data = public_libraries,
             lng = ~longitude,
             lat = ~latitude,
             color = ~pal_system(overagency),
             fill = ~pal_system(overagency),
             opacity = 1,
             fillOpacity = 1,
             radius = 100,
             label = ~facname,
             popup = ~address#,
             #group = "Broadband"
             ) %>% 
    # addCircles(data = public_libraries,
    #          lng = ~longitude,
    #          lat = ~latitude,
    #          color = ~pal_system(overagency),
    #          fill = ~pal_system(overagency),
    #          opacity = 1,
    #          fillOpacity = 1,
    #          radius = 100,
    #          label = ~facname,
    #          popup = ~address,
    #          group = "Income"
    #          ) %>% 
  addLegend(position = "topleft",
            pal = pal_system, 
            values = public_libraries$overagency,
            title = "Library System") %>% 
  addLegend(position = "topleft",
            pal = pal_broadband, 
            values =  pumas_2010_geojson$broadband_adoption_quartile,
            opacity = .7,
            title = "Broadband Access",
            group = "Broadband") %>% 
  addLegend(position = "topleft",
            pal = pal_income, 
            values =  pumas_2010_geojson$median_income_category,
            opacity = .7,
            title = "Median household income",
            group = "Income") %>% 
  addLayersControl(
            baseGroups = c("Broadband","Income"),
            options = layersControlOptions(collapsed = F)#,
            #overlayGroups = c("Branches")
  ) 


libraries_map

```

The htmlwidgets package lets you save this map as an html file, which can then be opened in a browser or embedded in a web page

```{r}
htmlwidgets::saveWidget(libraries_map,
                        file = "libraries_map.html",
                        title = "NYC Libraries:  Broadband Access and Median Income")

```

## Other resources

-   [`tidycensus`](https://walker-data.com/tidycensus/) package: amazing R interface for Census/ACS API, including geometries

    -   [Analyzing Census Data: Methods Maps and Models in R](https://walker-data.com/census-r/)

-   [`tigris`](https://github.com/walkerke/tigris) package: more Census geometries

-   [`rnaturalearth`](https://docs.ropensci.org/rnaturalearth/): world map data

-   [`tmap`](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html): Alternative package for tmapping in R, with ggplot-like syntax - some find this the best way to get started

-   [`mapboxapi`](https://walker-data.com/mapboxapi/) package: more basemap options, geocoding services, isochrones (requires setting up a mapbox account)

-   Geocoding

    -   Within NYC, [Geosupport](https://www1.nyc.gov/site/planning/data-maps/open-data/dwn-gde-home.page)

    -   [Urban Institute overview of other tools](https://urban-institute.medium.com/choosing-a-geocoder-for-the-urban-institute-86192f656c5f)

-   [Geocomputation with R](https://geocompr.robinlovelace.net/index.html)
